### 2019-05-24 chapter_00_open_science.tex
Correct Python section and start the workflow section

\item \textbf{Multi-paradigm.} Supports imperative, object-oriented, functional
and aspect-oriented programming.

@@

\item Scripting layer for applications (for example Paraview, Visit, QGIS, Blender, ...)

@@

Since the standard Python interpreter does not do any proper
compilation, Python code for some CPU bounded tasks can be too slow. Thus, one
has to use special tools to get optimized machine instructions to speedup
performance-critical code.

\paragraphbf{Ahead Of Time compilation} Python is primarily a ``glue''
language, and a common solution is to use code written in C, C++ or Fortran.
%
It is possible, without a compilation step, to call functions in C libraries
through pure python code using \pack{ctypes} standard module or the \pack{cffi}
package.
%
Libraries with Python bindings can also be used directly (for e.g., wrapped
using SWIG or Boost.Python) or one has to write extensions, i.e.\ Python
modules written in C or C++ using the CPython API that can be imported directly
from Python like any Python module.
%
However, to rely on such techniques to call C libraries would inhibit
development of extending functionalities, as they have a steeper learning
curve compared to Python.
%
Nevertheless, this strategy has given rise to the base modules of the scientific
Python stack, namely \Numpy, \Scipy and \pack{Matplotlib}.

Python developers do not actually have to write the compiled extensions in C,
C++ or Fortran, since there are tools to generate them automatically from
Python code (Pythran) or from Cython code. Cython \citep{behnel_cython2011} is a
smooth blend of two languages with possibility to add type declarations like in
C/C++, but with a syntax similar to Python.

Pythran \citep{guelton2015pythran, guelton2018pythran} is a quite recent Python
compiler which gives impressive results. It creates compiled extensions from pure
Python code with simple type annotations written as comments.
%
The resulting extensions are usually as fast as Fortran or C++ written by
non-specialists \citep[see, for example, benchmarks for the packages
\fluidpack{fft} and \fluidpack{sim},][]{fluidfft, fluidsim}.
%
They are created with a very interesting two-step compilations: first the code
is optimized at the Python level and then an automatically produced C++ code is
properly compiled.
%
This two-step compilation provides room for very clever optimizations as
demonstrated, for example, in
\href{http://serge-sans-paille.github.io/pythran-stories/%
being-more-than-a-translator.html}{%
this blog post by the main developer of Pythran}.
%
Pythran supports OpenMP pragma \citep{guelton2013compiling} and can use modern
vectorization with SIMD instructions \citep{guelton2014exploring}. Pythran
understands both Matlab-like vectorized code and C-like code with explicit loops
\citep{guelton2018pythran}.

We present in \citet{fluidfft, fluidsim} examples of
highly efficient Python codes, optimized using a combination of Cython and
Pythran extensions.

\paragraphbf{Just In Time (JIT) compilation} Another way to obtain optimized
machine instructions is by using JIT compilation, i.e.\ by compiling only the
critical code at run time.
%
This strategy can yield good results with other languages as for example Matlab
or Julia.
%
Pypy, an alternative interpreter written in Python, has a JIT compiler. However,
Pypy is not widely used for scientific applications mainly because of
compatibility problems with the extensions written for CPython.
%
However this could change since a recent version of Pypy (6.0) supports the main
packages of the Python scientific stack (\Numpy, \Scipy, \pack{Matplotlib},
\pack{Pandas}, etc.).

Adding a JIT to the interpreter CPython has been notoriously difficult (see
\href{https://faster-cpython.readthedocs.io/}{faster-cpython}). One of the
problem seems to be the C API provided in CPython and used by many extensions.
%
Python being first a glue language, a faster CPython with a JIT is actually not
so necessary for many purposes. Therefore, not as much money and work has been
put in accelerating Python than for example for Java and JavaScript.

Another strategy is to add a JIT to CPython through an external package
(\pack{Numba}) and to only compile the critical functions.  Moreover,
\pack{Numba} is particularly interesting because it can take advantage of the
GPU\footnote{See for example
\url{https://devblogs.nvidia.com/parallelforall/seven-things-numba/}}.

\paragraphbf{No type-checking} In contrast to languages using Ahead Of Time
compilation, there is no type-checking in Python (and more generally in many
dynamical languages).  Type-checking is useful since it can avoid simple bugs.
However, it is not a good practice even with static languages to rely only on
type-checking to look for bugs in a code.
%
It is known that a rigorous set of unit tests is also needed to obtain more
reliable codes.
%
Dynamical languages rely a lot on unit tests so it is very important to put at
least a little bit of time and energy to write a decent battery of tests.
%
We have to mention the young and promising project
\href{http://mypy-lang.org/}{Mypy}\footnote{Guido van Rossum, the author and
``Benevolent Dictator For Life'' of Python, is involved in the project Mypy.},
which allows for Optional Static Typing and thus type-checking for
Python\footnote{We plan to add type hitting in the most important FluidDyn modules
to investigate how we can take advantage of type-checking with Mypy.}.

\paragraphbf{A lively, huge and thus complicated ecosystem} As already
mentioned, Python is a versatile language. For many applications, the standard
library is insufficient and one needs external packages.
%
% For most applications there are several projects usable through Python.
Most of the time there are several Python packages for a certain application.
A user/developer has to make choices regarding the tools to use.  Especially
for the beginners who are not used to the open-source realm, it can be
difficult to understand a ``landscape of projects'' for a particular
application and to make good technological choices.

It is important to get a good introduction on how to work with the open-source
dynamics and the scientific Python ecosystem.  It can also be very useful to ask
to more advanced users and attend Python courses or workshops\footnote{This could
be regular academic courses, MOOCs or workshops like ones organized by
\href{https://software-carpentry.org/}{Software Carpentry}.}.
%
Of course, such abundance of projects can also be seen as an advantage.

\paragraphbf{Absence of multicore computational parallelism using threads}
Threads behave like light subprograms. In principle, they can use the different
cores of the CPU at the same time but the CPython interpreter forbids that they
interpret python code at the same time\footnote{See
\href{https://opensource.com/article/17/4/grok-gil}{this blog article on ``How
to grok the GIL''} and
\href{https://faster-cpython.readthedocs.io/cpython37.html}{this technical
presentation by a CPython core developer}.}.
%
This limitation comes from an important detail in the implementation of the
CPython interpreter: a Global Interpreter Lock (GIL) to prevent race conditions
that could corrupt data.
%
It seems that the GIL greatly simplies the implementation of CPython and that it
is very difficult to remove it while keeping other nice technical properties of
CPython intact\footnote{See \url{https://wiki.python.org/moin/GlobalInterpreterLock}.}.
Note that Pypy also uses a GIL, while two other implementations of the Python
language, Jython and IronPython, do not have this limitation.

At any rate, threads using the interpreter do not use the CPU at the same time
so it is not possible to use threads to do multicore computational parallelism
with pure python code to accelerate CPU bounded tasks.
%
Note however, that threads can be (and are widely) used in Python for
concurrency, i.e.\ to perform I/O tasks (which does not need CPU) concurrently.
%
To do proper computational parallelism for CPU bounded tasks with the GIL, one has
to use other strategies: extensions (fine grain parallelism), multiprocessing
(coarse grain parallelism) and inter-process communication, for example, with MPI
(with mpi4py) or ZeroMQ.

\subsection{Rich landscape of open-source languages useful for science}

@@

Fortran remains widely used for computing codes, at least in fluid mechanics.
C++ is used in many recent developments. Its recent improvements (C++11, C++14)
make it a modern and very useful language.
%
JavaScript, Java, Scala, Smalltalk, Haskell, R, Julia, Perl and Lua are used for
some scientific codes. Go and Rust are quite young languages and are not (yet)
widely used for scientific applications.

It is of course very difficult to predict which technologies will be widely
used in even the near future (10 or 20 years for example).
%
From what we see, it seems that the idea of ``one language to do everything for
science'' will not succeed, at least soon.  This was maybe the project of Julia
creators, but despite the qualities of this language, it also has
limitations\footnote{For example C-like include instead of Python-like
import. From our point of view, a Pythonic flavor of Julia would be very
interesting.}.
%
It seems that a very important aspect in the future will be interoperability
between coexisting tools\footnote{See for example the cross-language development
platform \href{https://arrow.apache.org/}{Apache Arrow}.}.

So what about Python?  As already mentioned, it is today the language with the
strongest dynamics in science, data analysis and much more.  Several students
learn and will learn it. Many scientists in the future would, hopefully, be
able to appreciate these strengths and favour the Pythonic approaches.  With
such dynamics today, we can anticipate that the Python world will be able to
embrace new technological trends and to overcome some of its current
limitations.
%
It is reasonable to predict that Python will continue to shine as a language
(i) particularly good for human communication and to write ideas, (ii) for fast
prototyping and (iii) for interoperability with code written in other languages
(see for example the recent projects \href{https://cppyy.readthedocs.io}{cppyy}
and \href{https://github.com/pybind/pybind11/}{pybind11}, or
\href{https://github.com/PyO3}{the possibility to interact with Rust code}).

\paragraphbf{Summary: Python, a versatile Swiss Army Knife for the
scientists}

Python has become a great versatile tool for scientists.  The scientific
ecosystem of Python is nowadays really impressive.  The scientific Python
community is strong and quickly growing.
%
A scientist with good skills in Python can do most of what she/he needs to do in
programming (which does not mean that Python is the best language to do
anything!).
%
In contrast, no skill in Python is a real disadvantage for many tasks and for
employability.
%
Therefore, in most fields today, if a scientist has to acquire good skills in
one language, it is in Python.

Considering what we have presented on Python, we think that it is reasonable to
bet that it will become one of the \textit{lingua franca} in science and in
fluid dynamics.  This explains why we have chosen it to be the main language
for the project FluidDyn.

@@

% remark Julien Salort "It is well known that": bad
% pa: common sense says ?
% It is well known 
It is not surprising
% av: "It is not surprising"
%
that short-term efficiency and long-term efficiency are
sometimes incompatible.
%
Quick and dirty scripts can be efficient in the short term, but such a workflow
has a negative impact on a longer time scale.

Similarly, we can also differentiate productivity at different scales of
collaboration.
%
An individual can be very efficient with a particular tool, but the same tool
can be very inefficient to collaborate even with co-workers.
%
A group can be very efficient with a code which is closed-source.  If this
strategy can be efficient for the group (at least in the short term), it leads
to waste for the community. Other groups will need to develop codes with the
same features.  Good ideas will be spread in different codes and since the
number of users and developers are smaller for each of these codes, the code
quality will not improve as fast as it would with an open-source strategy. In a
nutshell, as this proverb says, \textit{``If you want to go fast, go alone. But
if you want to go far, go together.''}

Thus, a community using open-source methods can be efficient to co-develop its
tools. However, some of its members have to spend more energy to bear this
dynamics, which implies that they somehow sacrifice their individual efficiency.
We see that short-term efficiency, long-term efficiency and productivity at
different levels can often be incompatible and that we have to consider these
conflicts when choosing between different technologies.

@@

We study fluid mechanics via laboratory experiments, in situ measurements,
analytic computation, numerical simulations and data processing.
%
Nowadays, programming is involved in all methods, and thus it is difficult to do
fluid mechanics without software development.
%
However, the level in software engineering is on average very low in the
community.
%
Even today, it is quite commonplace to start a PhD without any serious training in
GNU/Linux and modern programming tools.
%
A large majority of scientists and technicians are also unaware of the new
challenges and opportunities of open-source.

>>>>

\item \textbf{Multi-paradigm.} Imperative, object-oriented, functional
and aspect-oriented programming are supported.

@@

\item Scripting layer for applications (for example Paraview, Visit, QGIS,
	Blender)

@@

In the context of scientific computing there are some common issues that Python
developers would face. The issues are listed below and the solutions are
specified briefly. The solutions are explained in further detail in subsequent
paragraphs.

\begin{itemize}

\item \textbf{Performance in CPU bounded tasks.} Since the standard Python
	interpreter does not do any proper compilation, pure Python code for
	some CPU bounded tasks can be too slow. Thus, one has to avoid writing
	loops as much as possible and use vectorized array operations as much
	as possible, or call \emph{foreign functions}, or use special tools to
	build extensions using \emph{Ahead Of Time} or \emph{Just In Time}
	compilation to speedup performance-critical code.

\item \textbf{Absence of type checking.} In contrast to compiled languages,
	there is no type-checking in Python (and more generally in many
	interpreted languages). Types are respected however, and type errors,
	if any, would occur during runtime.
	\emph{Type hints} and \emph{static type checking} can solve this issue.

\item \textbf{Concurrent, but not parallel multi-threading.} Threads behave
	like light sub-programs. In principle, they can use the different cores
	of the CPU at the same time, but the CPython interpreter forbid threads
	to interpret python code at the same time\footnote{See
	\href{https://opensource.com/article/17/4/grok-gil}{this blog article
on ``How to grok the GIL''} and
\href{https://faster-cpython.readthedocs.io/cpython37.html}{this technical
presentation by a CPython core developer}.}. In other words, threads can
coexist concurrently, by not execute in parallel.
%
This limitation comes from an important detail in the implementation of the
CPython interpreter: a Global Interpreter Lock (GIL) to prevent race conditions
that could corrupt data.
%
GIL greatly simplifies the implementation of CPython and that
it is very difficult to remove it while keeping other nice technical properties
of CPython intact\footnote{See
\url{https://wiki.python.org/moin/GlobalInterpreterLock}. Note that PyPy
also uses a GIL, while two other implementations of the Python language, Jython
and IronPython, do not have this limitation}.

\item \textbf{Lively, and thus, complicated ecosystem of packages.} For many
	applications, the standard library is insufficient and one needs
	external packages. Sometimes there are multiple options and the
	user/developer has to make choices regarding the tools to use.
	Especially for the beginners who are not used to the open-source realm,
	it can be confusing prospect to make good technological choices.

\subsubsection{Foreign functions}
%
Python is primarily a ``glue'' language, and a common solution to achieve
performance is to use existing code written in C, C++ or Fortran.
%
It is possible, without a compilation step, to call functions in C libraries
through pure python code using \pack{ctypes} standard module or the \pack{cffi}
package.
%
Similarly \href{https://cppyy.readthedocs.io}{cppyy} can interpret C++ code.
Libraries with Python bindings can also be used directly (for e.g., wrapped
using SWIG or Boost.Python) or one has to write extensions, i.e.\ Python
modules written in C or C++ using the CPython API that can be imported directly
from Python like any Python module.
%
However, to rely on such techniques to call C libraries would inhibit
development of extending functionalities, as they have a steeper learning
curve compared to Python.
%
Nevertheless, this strategy has given rise to the base modules of the scientific
Python stack, namely \Numpy, \Scipy and \pack{Matplotlib}.

\subsubsection{Ahead Of Time compilation}
Ahead Of Time compilation is process of generating compiled extensions which
can be dynamically executed by Python.
%
Python developers do not actually have to write the extensions in C, C++ or
Fortran, since there are tools to generate them automatically from Python code
(Pythran) or from Cython code. Cython \citep{behnel_cython2011} is a smooth
blend of two languages with possibility to add type declarations like in C/C++,
but with a syntax similar to Python.
%
There are also other projects to create extensions in C++ with
\href{https://github.com/pybind/pybind11/}{pybind11}, and in Rust with
\href{https://github.com/PyO3}{PyO3}.

Pythran \citep{guelton2015pythran, guelton2018pythran} is a quite recent Python
compiler which gives impressive results. It creates compiled extensions from pure
Python code with simple type annotations written as comments.
%
The resulting extensions are usually as fast as Fortran or C++ written by
non-specialists \citep[see, for example, benchmarks for the packages
\fluidpack{fft} and \fluidpack{sim},][]{fluidfft, fluidsim}.
%
They are created with a very interesting two-step compilations: first the code
is optimized at the Python level and then an automatically produced C++ code is
properly compiled.
%
This two-step compilation provides room for very clever optimizations
\footnote{As demonstrated in this blog post by the main developer of
Pythran \url{http://serge-sans-paille.github.io/pythran-stories/%
being-more-than-a-translator.html}}.
%
Pythran supports OpenMP pragma \citep{guelton2013compiling} and can use modern
vectorization with SIMD instructions \citep{guelton2014exploring}. Pythran
understands both Matlab-like vectorized code and C-like code with explicit loops
\citep{guelton2018pythran}.
%
We present in \citet{fluidfft, fluidsim} examples of
highly efficient Python codes, optimized using a combination of Cython and
Pythran extensions.

\subsubsection{Just In Time (JIT) compilation}
Another way to obtain optimized machine instructions is by using JIT
compilation, i.e.\ by compiling only the critical code at run time.
%
This strategy can yield good results with other languages as for example Matlab
or Julia.
%
PyPy, an alternative interpreter written in Python, has a JIT compiler. However,
PyPy is not widely used for scientific applications mainly because of
compatibility problems with the extensions written for CPython.
%
However this could change since recent versions, starting with PyPy 6.0,
supports the main packages of the Python scientific stack (\Numpy, \Scipy,
\pack{Matplotlib}, \pack{Pandas}, etc.).

Adding a JIT to the interpreter CPython has been notoriously difficult
\footnote{\url{https://faster-cpython.readthedocs.io/}}. One of the
problem seems to be the C API provided in CPython and used by many extensions.
%
A faster CPython with a JIT is actually not so necessary for many purposes,
since Python extensions are enough. Therefore, not as much money and work
has been put in accelerating Python than for example for Java and JavaScript.

Another strategy is to add a JIT to CPython through an external package
(\pack{Numba}).
%
\pack{Numba} is particularly interesting because it can take advantage of the
GPU\footnote{See for example
\url{https://devblogs.nvidia.com/parallelforall/seven-things-numba/}}.

\subsubsection{Type hints and static type checking}
%
Type-checking is useful since it can avoid simple bugs.
\fnref{https://docs.python.org/3/whatsnew/3.5.html\#whatsnew-pep-484}{PEP-484}
introduced a syntax and a related \codeinline{typing} module to add
type-hints, typically for function parameters and return types.
%
Static type checkers like \href{http://mypy-lang.org/}{Mypy} and
\href{https://github.com/google/pytype}{pytype}
%
\footnote{We plan to add type hinting in the most important FluidDyn modules to
investigate how we can take advantage of type-checking with Mypy.}
%
allows to process type-hints and report inconsistencies in the code.

However, it is not a good practice even with static languages to rely only on
type-checking to look for bugs in a code.
%
It is known that a rigorous set of unit tests is also needed to obtain more
reliable codes.
%
Dynamical languages rely a lot on unit tests so it is very important to put at
least a little bit of time and energy to write a decent battery of tests.

\subsubsection{Achieving computational parallelism}

At any rate, threads using the interpreter do not use the CPU at the same time
so it is not possible to use threads to do multi-core computational parallelism
with pure python code to accelerate CPU bounded tasks.
%
Note however, that threads can be (and are widely) used in Python for
concurrency, i.e.\ to perform I/O tasks (which does not need CPU) concurrently.
%
To do proper computational parallelism for CPU bounded tasks with the GIL, one
has to use other strategies: extensions (fine grain parallelism),
multiprocessing (coarse grain parallelism) and inter-process communication, for
example, with MPI (with mpi4py) or ZeroMQ.

\subsubsection{Choosing the right package}
%
For a beginner in Python development, it is important to get a good
introduction on how to work with the open-source workflows and understanding of
the scientific Python stack.
%
It can also be very useful to ask to more advanced users and attend Python
courses or workshops\footnote{This could be regular academic courses, MOOCs or
workshops like ones organized by
\href{https://software-carpentry.org/}{Software Carpentry}.}.
%
Of course, such abundance of projects can also be seen as an advantage.

\subsection{The state of languages for scientific computing}

@@

Fortran remains widely used for computing codes, at least in fluid mechanics.
C++ is used in many recent developments. Its recent improvements (C++11, C++14)
make it a modern and very useful language.
%
JavaScript, Java, Scala, Smalltalk, Haskell, R, Julia, Perl and Lua are used for
some scientific codes. Go and Rust are quite young languages with emphasis on
memory-safety and are not (yet) widely used for scientific applications.
%
It is, of course, very difficult to predict which technologies will be widely
used in even the near future (10 or 20 years for example).
%
It seems that the idea of ``one language to do everything for science'' will
not succeed, at least not in the near future. 
%
The limited scope of the language is likely the reason why the community
around Fortran and Julia are not widespread outside academia.
%
It seems that a very important aspect in the future will be interoperability
between coexisting tools\footnote{See for example the cross-language development
platform for in-memory data, \href{https://arrow.apache.org/}{Apache Arrow}.}.
%
Knowledge of Python is also an advantage in terms of employability these days.
%
Considering what we have presented on Python, we think that it is reasonable to
bet that it will become one of the \textit{lingua franca} in science and in
fluid dynamics.
%
We have, therefore, chosen it to be the main language for the project FluidDyn.

@@

% remark Julien Salort "It is well known that": bad pa: common sense says ?  It
% is well known 
It is not surprising
% av: "It is not surprising"
%
that short-term efficiency and long-term efficiency are sometimes incompatible.
%
Quick and dirty scripts can be efficient in the short term, but has a negative
impact on a longer time scale.
%
Similarly, we can also differentiate productivity at different levels of
collaboration.
%
An individual can be very efficient with a particular tool, but the same tool
can be very inefficient when used in a collaboration.
%
A group can be very efficient with a code which is closed-source. Although such
a strategy can be efficient for the group (at least in the short term), it may
lead to waste for the community. Other groups will need to develop codes with
the same features.
%
Good ideas will be spread in separate codes and since the number of users and
developers is small for each of these codes, the code quality will not
improve as fast as it would with an open-source strategy. In a nutshell, as
this proverb says, \textit{``If you want to go fast, go alone. But if you want
to go far, go together.''}

Thus, a community using open-source methods can be efficient to co-develop its
tools. However, some of its members have to spend more energy to bear this
dynamics, at the risk of sacrificing their individual efficiency.
We see that short-term efficiency, long-term efficiency and productivity at
different levels can often be incompatible and that we have to consider these
conflicts when choosing between different technologies.

@@

We study fluid mechanics via laboratory experiments, in situ measurements,
analytic tools, numerical simulations and data processing.
%
Nowadays, programming is involved in all methods, and thus it is difficult to do
fluid mechanics without software development.
%
However, the level in software engineering is on average very low in the
community.
%
Even today, it is quite commonplace to start a Ph.D. without any serious
training in GNU/Linux and modern programming tools.
%
A large majority of scientists and technicians are also unaware of the new
challenges and opportunities of open-source.
